---
title: "DAG learning with R package `clrdag`"
author: "Chunlin Li"
institute: School of Statistics, University of Minnesota
date: "April 29th, 2019"
output: beamer_presentation
header-includes:
- \usepackage{amssymb,amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## DAG learning 

Structual equations model:
    $$ X_k = \sum_{j\in \text{pa}_k} A_{jk} X_j + \varepsilon_k, \quad k = 1,\ldots,p, $$
    where $A_{jk} \neq 0$ iff $j \in \text{pa}_k$ and $\varepsilon_k \overset{iid}{\sim} N(0,\sigma^2)$.

\vskip .5cm
    
The iid errors $\varepsilon_k$ implies $A_{p\times p}$ is a (weighted) adjacency matrix of a \textcolor{red}{directed acyclic graph}. Denote $A\in\mathbb D$. 

\vskip .5cm

Assume $\mathbf X_i = (X_{i1},\ldots,X_{ip})$; $i=1,\ldots,n$ are iid observations. Let $\mathbb X_{n\times p} = (\mathbf X_1',\ldots,\mathbf X_n')'$.


\vskip .5cm

\textcolor{red}{Goal} Estimate and make inference on $A$.

## Likelihood: original

Constrained MLE: \textcolor{red}{difficult!} 
$$ \max_{\textcolor{cyan}{A\in \mathbb D}} \quad  \ell(A) = \frac{1}{2} \| \mathbb X - \mathbb X A \|_F^2 + \textcolor{blue}{\mu\|A\|_{off,0}}.$$ 
Relaxation? 






## Likelihood: relaxation

Iterative convex relaxation (DC)\footnote{The details are included in the paper and R package vignette.}:
    
\vskip .5cm

\quad At $(t+1)$th step, $B^{[t]} = H_{\tau}(A^{[t]})$; $H_{\tau}$ is hard threshold.
    \begin{equation*}
        \begin{split}
            (A^{[t+1]},\Lambda^{[t+1]}) = \arg\max \frac{1}{2} \| \mathbb X - \mathbb X A \|_F^2 + \textcolor{blue}{\mu \tau^{-1} \| B^{[t]} \circ A \|_{off,1}} \\
            \text{s.t.} \ \textcolor{cyan}{\lambda_{jk} + \mathbb I(i\neq k) - \lambda_{ik} \geq \tau^{-1} |A_{ij}|B^{[t]}_{ij} + (1 - B^{[t]}_{ij}),}\\
             \textcolor{cyan}{i,j,k = 1,\ldots,p;\quad i\neq j.}
        \end{split}
    \end{equation*}

\quad Iterate until objective converges.



## Computation

Our implementation is based ADMM.

\vskip .5cm

\textcolor{red}{Algorithm} 
Initiate $(A^{[0]}, \Lambda^{[0]})$ satisfying DAG constraint. 
Set $B^{[0]} = H_{\tau}(A^{[0]})$. 

Set the optimization accuracy $\epsilon > 0$, for $t=1,\ldots,\infty$
\begin{enumerate}
\item Compute $(A^{[t]},\Lambda^{[t]})$ by ADMM. Set $B^{[t]} = H_{\tau}(A^{[t]})$. 
\item If $B^{[t]}$ has a cycle, for each $|A^{[t]}_{ij}|>0$ in increasing order, if $(i,j)$ is in a cycle, 
\[ A^{[t]}_{ij} \leftarrow 0, \quad B^{[t]}_{ij}\leftarrow 0. \]
\end{enumerate}





## `clrdag`

C++ and R (`Rcpp`) with OpenMP:
\begin{itemize}
\item Friendly API.
\item Highly optimized linear algebra libraries (`Armadillo`).
\end{itemize}

\vskip .5cm

Do one thing and do it well: `MLEdag`
\begin{itemize}
\item Compute MLE for DAG.
\item Make likelihood inference on DAG edges.
\end{itemize}


## A toy example: setup

\begin{itemize}
  \item $p = 50, n = 1000$
  \item Random DAG $A$ with connection probability $1/p$
  \item Nonzero $A_{ij}$ are assigned Unif$[-1,-0.7]\cup[0.7,1]$
  \item $\varepsilon_k \overset{iid}{\sim} N(0,1)$
\end{itemize}

## A toy example: $\|A\|_0 = 43$

```{r, fig.height=4, fig.width=4}
library(lattice)
p <- 50; n <- 1000; sparsity <- 2/p
set.seed(2018)
idx <- sample(1:p)
set.seed(2019)
A <- matrix(rbinom(p*p,1,sparsity)*sign(runif(p*p,min=-1,max=1))*runif(p*p,min = 0.7, max = 1),p,p)
A[upper.tri(A, diag = TRUE)] <- 0
A <- A[idx,idx]
A_plot <- levelplot(A, col.regions = colorRampPalette(c("blue", "white", "red"))(1e3),at=seq(min(-1.2), max(1.2), length.out=50))
A_plot
```


## A toy example: code

```{r}
X <- matrix(rnorm(n*p), n, p) %*% t(solve(diag(p) - A))
```

```{r, echo=TRUE}
library(clrdag)
t <- proc.time()
out <- MLEdag(X=X,tau=0.3,mu=1,rho=1.2)
proc.time() - t
```

## A toy example: result

$\|\hat{A} - A \|_F^2$: 
```{r, echo=TRUE}
sum((out$A - A)^2)
```

\vskip .5cm

$\text{Hamming}(\mathcal G_{\hat A},\mathcal G_A)$:
```{r, echo=TRUE}
sum(abs((out$A != 0) - (A!=0)))
```


## Summary

URL: [`https://github.umn.edu/li000007/clrdag`](https://github.umn.edu/li000007/clrdag)

\vskip .5cm

Due to time limit, the inference part is not presented.

\vskip .5cm

TODOs: 
\begin{itemize}
        \item \textcolor{red}{Cross-validation function}, diagnostic plots.
        \item Publish on CRAN.
\end{itemize}



## Thank you

References
\begin{itemize}
    \item Li, C., Shen, X., \& Pan, W. (2019). 
    Likelihood ratio tests of a large directed acyclic graph. 
    Submitted. 

  \item Yuan, Y., Shen, X., Pan, W., \& Wang, Z. (2018). 
  Constrained likelihood for reconstructing a directed acyclic Gaussian graph. 
  \emph{Biometrika}, 106(1), 109-125.

\end{itemize}
















